---
title: "Final Project Proposal + Data"
author: "Stolie Erickson, Mira Shlimenzon, Joseph Konat, Casey Avila"
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
    number-sections: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
options(scipen = 999)
co2 <- read_csv("co2.csv")
ipp <- read_csv("ipp.csv")
```

## Data Description

The two variables that we are interested in are the average daily income per capita (independent variable) and how it is related to CO2 emissions per capita (response variable).

### Income Dataset

The income dataset contains information regarding each country's average daily household income based on a 2011 survey. These are measured in constant international dollars (PPP), converting the dollars to amounts that equalize the purchasing power across different currencies/regions. There does not appear to be NA values or a number that is replacing missing data, although our data cleaning will reveal more information regarding this. 
On Gapminder, this dataset is titled "Average daily income per capita, PPP (constant 2011 international $)".

### CO2 Dataset

The CO2 data set describes the consumption based CO2 emissions in million tonnes of CO2 per capita. These values are based on the Global Carbon Project 2021. On Gapminder, this dataset is titled "CO2 emissions per capita (Consumption based)".

## Hypothesis

According to some [research](https://www.cgdev.org/media/who-caused-climate-change-historically#:~:text=Developed%20Countries%20Are%20Responsible%20for,Global%20Development%20%7C%20Ideas%20to%20Action), we anticipate that more economically developed countries (those with higher average incomes per capita) will be correlated with higher CO2 emissions compared to countries with lower average incomes per capita.

## Data Cleaning

### CO2 Dataset

Upon initially looking through the CO2 data, we noticed that less economically developed countries in earlier years have repeated emission values for extended periods of time. We believe that it is unreasonable for CO2 emissions per capita to remain exactly the same for 30+ years for several rows. Under the assumption that predictive modeling was used to fill in data which was unknown dating back to 1800, we will be taking steps to clean it. In order to prevent our data from skewing, we are creating a function that will take in each row of the data set, and replace the emission values with NA’s until the emission initially changes. For example, since Australia has an emission value of 0 from 1800 to 1859, then changes in 1860 to .33, our function will sequence through the row and replace all of the 0’s with NA until 1859 (it will keep only the last 0 before the change) then the rest of the row will remain unchanged. Additionally, we noticed how the minus signs of the CO2 emissions per capita were not actual minus signs, so we replaced them accordingly. 

```{r}
clean_init_repeats <- function(df) {
  for (i in 1:nrow(df)) {
    row <- df[i,]
    for (j in 2:(length(row)- 1)) {
      if (row[j] == row[j + 1]) {
        row[j] <- NA
      } else {
        break;
      }
    }
    
    df[i,] <- row
  }
  
  return(df)
}

co2 <- co2 %>%
  clean_init_repeats()

```

### Income Dataset

This dataset suffered from a similar problem to that of the CO2 dataset, so we can
re-use our cleaning function on `ipp`.

```{r}
ipp <- ipp %>%
  clean_init_repeats()
```

## Final Cleanup and Pivoting

### CO2 Dataset

Our CO2 dataset also included negative values, which were encoded as strings due
to them using a "−", rather than "-" (note the slight difference). This was fixed
by replacing the dash with a traditional hyphen so that `as.numeric` could interpret
the string, and then converting the values.

```{r}
co2_long <- co2 %>%
  mutate(across(`1800`:`2020`, 
                ~ as.numeric(str_replace(.x, "−", "-")))) %>%
  pivot_longer(cols = `1800`:`2020`,
               names_to = "year",
               values_to = "emissions")
```

### Income Dataset

This data set included values from the future (2023 - 2050), which were predicted based
off of GDP growth rates for a given country. We removed all values after 2020 in
order to match our CO2 dataset and make joining simpler. We also wanted to avoid any predictive modeling that the income dataset may have used in order to obtain values for years that have not already occurred.

```{r}
ipp_long <- ipp %>%
   select(country:`2020`) %>%
   pivot_longer(cols = `1800`:`2020`,
               names_to = "year",
               values_to = "average_income_per_capita")
```

## Joining the Data

While merging our data, we wanted to ensure that both the explanatory and response variables remained grouped by both the country for which the data was collected and the year. This will ensure that if we want to analyze specific countries then we can group by that aspect, similarly with the year. This will allow us to add details to our visualizations, if desired. 

We decided to drop all of the NA values in the merged data set. This is because with regards to correlation and linear regression, it is generally recommended to remove rows with missing values. When there are missing values in either the predictor or the response variables it can create inconsistencies in the data. This could potentially lead to biased or inaccurate results. Although lm and ggplot automatically drop the NA values, if we plan on doing further summary analysis we want to use the data that is demonstrated in our models.
```{r join}
merged_data <- inner_join(co2_long, ipp_long, by = c("country", "year")) %>%
  drop_na()
```

##Linear Regression

##Data Visualization

*The relationship between the two variables*
```{r}
merged_data |>
  ggplot(mapping = aes(x = average_income_per_capita, y = emissions)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  labs(x = "Average Income Per Capita (PPP)", y = "CO2 Emissions Per Capita (million tonnes)", title = "Relationship between Average Income and CO2 Emissions Per Capita")
```
*how this relationship has changed over time*
*Need help fixing the year axis, maybe make it so the years are every 20-50 years or something?!*
```{r}
merged_data |>
  mutate(year = as.factor(year)) |>
  ggplot(mapping = aes(x = year, y = average_income_per_capita, color = emissions)) +
  geom_line() +
  scale_x_discrete(breaks = merged_data$year[seq(1, length(merged_data$year), by = 20)], labels = merged_data$year[seq(1, length(merged_data$year), by = 20)])
  labs(x = "Year", y = "Average Income Per Capita (PPP)", title = "Relationship between Average Income and CO2 Emissions Per Capita") +
  scale_color_gradient(low = "blue", high = "red", name = "Emissions (million tonnes)")
```

##Linear Regression Model
*this is how the model will look if we use all of the data for the regression. She says we can choose to do this or make it one x and one y value per country so we just need to decide*
```{r}
linear_data <- merged_data |>
  filter(year %in% c(1990: 2020)) |>
  group_by(country) |>
  summarize(average_emission = mean(emissions), average_income = mean(average_income_per_capita))
```


```{r}
Model <- lm(average_emission ~ average_income, data = linear_data)
summary(Model)
```

```{r}
#printing out the regression equation
coef <- coef(Model)
print(paste("The Regression Equation is: Emissions =", as.character(coef[1]), "+", as.character(coef[2]), "*average_income"))
```

Interpretation of the Intercept: For countries with an average income per capita of 0 constant international dollars, the estimated CO2 emissions per capita is .505. *do we have data points at 0? if not, this is not interpretable*

Interpretation of average_income_per_capita: Each increase in one constant international dollar is associated with an increase of .278 in CO2 emissions per capita for all countries similar to those in the study.

